Effective multi-body SU($N$)-symmetric interactions of ultracold fermionic atoms on a 3-D lattice

### 2019-02-04 multi_body_interactions/papers/theory/reviews_NJP/referee_1_response.tex

\documentclass[preprint]{revtex4-1}

% linking references
\usepackage{hyperref}
\hypersetup{
  breaklinks=true,
  colorlinks=true,
  linkcolor=blue,
  urlcolor=cyan,
}

% general physics / math packages and commands
\usepackage{physics,amsmath,amssymb,braket,dsfont,bm}
\renewcommand{\t}{\text} % text in math mode
\newcommand{\f}{\dfrac} % shorthand for fractions
\newcommand{\p}[1]{\left(#1\right)} % parenthesis
\renewcommand{\sp}[1]{\left[#1\right]} % square parenthesis
\renewcommand{\set}[1]{\left\{#1\right\}} % curly parenthesis
\newcommand{\bk}{\braket} % shorthand for braket

\newcommand{\g}{\text{g}}
\newcommand{\e}{\text{e}}
\newcommand{\x}{\text{x}}
\newcommand{\y}{\text{y}}
\newcommand{\z}{\text{z}}

\renewcommand{\c}{\hat{c}}
\newcommand{\n}{\hat{n}}

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\K}{\mathcal{K}}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\1}{\mathds{1}}

% inline lists
\usepackage[inline]{enumitem}
\setlist[enumerate]{leftmargin=*} % nice margins for enumerate

% color definitions (used in a figure)
\usepackage{xcolor}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\green}[1]{\textcolor{green}{#1}}
\definecolor{lightblue}{RGB}{31,119,180}
\definecolor{orange}{RGB}{255,127,14}
\definecolor{green}{RGB}{44,160,44}
\definecolor{lightred}{RGB}{214,39,40}

% for strikeout text
% normalem included to prevent underlining titles in the bibliography
\usepackage[normalem]{ulem}

\begin{document}

\section*{Referee 1}

We thank the referee for taking the time to carefully read our
manuscript, and for providing thoughtful and constructive feedback.
Below, we address the comments and concerns brought up by the referee.
We hope that the referee finds our responses and revisions
satisfactory, deeming our manuscript acceptable for publication in New
Journal of Physics.

Excerpts from the referee reports are written in \blue{blue}, excerpts
from our original manuscript are written in \red{red}, and new text in
the revised version of our manuscript is written in \green{green}.
All page, line, and reference numbers generally refer to those in our
original manuscript.

\begin{enumerate}
\item Concerning:

  \blue{The authors study the effective on-site interactions for the
    description of several fermionic particles confined in a very deep
    optical lattice. The focus is on alkaline earth atoms, where the
    s-wave interaction strength is independent on the nuclear spin,
    and on all atoms in the lowest state of the trapping
    potential. The manuscript is an extended version on the
    theoretical tools applied in the comparison between experiment and
    theory of Ref. [34].  The analysis is very important and
    timely. However, in the present form, the manuscript has some
    significant short comings, which have to be improved before a
    decision about publication can be made.}

  \blue{One of the main points of criticism is that the manuscript is
    very poorly written and very difficult to access. A few examples
    are listed below:}

  We thank the referee for the succinct summary of our work, and for
  recognizing that our analysis is ``very important and timely''.  We
  hope that thew revisions outlined below have made our manuscript
  more accessible, and that we have addressed shortcomings identified
  by the reviewer to a satisfactory degree.

\item Concerning the comment:

  \blue{(i) The authors try to be rather general in their description
    such as title, abstract and introduction. During the description
    of the results, they start to apply several approximations, which
    are neither mentioned before or justified.  As an example: the
    authors claim to derive a general low energy theory for orbital
    and nuclear spin stats. However, in Eq.(10) the authors suddenly
    restrict the analysis to a low density of excitations in the ``e''
    orbital state. This restriction is never explained nor motivated.}

  \label{pt:general}

  The referee is correct that we do not ``derive a general low energy
  theory for orbital and nuclear spin states''.  In our abstract, for
  example, we merely claim that (page 1, line 27)

  \red{Motivated in particular by recent experiments ... we develop a
    low-energy effective theory ... which exhibits emergent multi-body
    SU($N$)-symmetric interactions ...}

  In our introduction, meanwhile, the very first sentence discussing
  the technical contributions of our manuscript (page 4, line 19)
  explicitly states that

  \red{In this work, we investigate the first experimental
    capabilities with ultracold fermionic AEAs to prepare high-density
    samples in a 3-D lattice with multiple occupation of individual
    lattice sites [34].}

  We similarly begin both our abstract and introduction with a
  discussion of specific experiments, and make frequent reference to
  experimental developments and capabilities.  We thus believe that
  our presentation is faithful to the technical content of our
  manuscript; we do not make any claim of having developed a
  completely general theory two-orbital multi-body SU($N$)-symmetric
  interactions.  We claim only to have developed a low-energy theory
  that (i) is relevant for current experiments, and (ii) exhibits
  multi-body SU($N$)-symmetric interactions.  For these reasons, we do
  indeed introduce several approximations even in the first paragraph
  of our theory overview (section II, page 5).  We believe that
  theoretical work motivated by specific experiments is generally
  justified in making approximations that are relevant to the
  parameter regimes of these experiments.

  Nonetheless, we fully recognize and agree with the referee on the
  importance of making the scope of our technical contributions clear
  (as the referee comments later in point \ref{pt:restrictions}).  The
  fact that the referee at any point believed we will present a
  general theory without applying several experimentally-motivated
  approximations clearly signifies that we need to be more explicit in
  our aims.  Consequently, we have modified the second sentence of our
  original abstract, from (page 1, line 27)

  \red{Motivated in particular by recent experiments ...}

  to

  \green{In order to analyze recent experiments ...}

  We have also clarified an important point in our introduction,
  replacing the following text (page 4, line 22)

  \red{Specifically, we consider the preparation of isolated few-body
    systems in the deep-lattice limit, ...}

  with

  \green{Specifically, we consider ground-state preparation of
    isolated few-body systems in the deep-lattice limit, ...}

  The point on ground-state preparation in the deep lattice limit is
  important for a restriction we make later concerning single
  occupation of nuclear spin states on individual lattice sites.  This
  restriction (discussed in point \ref{pt:single_nuclear} below)
  together with the experimentally-motivated approximations made in
  the first paragraph of the theory overview (section II, page 5)
  constitute all key restrictions of our theory.

  We hope that these changes, together with those discussed in the
  examples below (points \ref{pt:orbital_state} and
  \ref{pt:single_nuclear}), clarify the motivation and scope of our
  manuscript, such that a reader is not led to believe that we present
  a general theory without applying several well-motivated
  approximations that are relevant to current experiments.

\item Concerning the specific example:

  \blue{However, in Eq.(10) the authors suddenly restrict the analysis
    to a low density of excitations in the ``e'' orbital state. This
    restriction is never explained nor motivated.}

  \label{pt:orbital_state}

  It is true, as the referee points out, that in Eq.~(10) (page 8) we
  present an effective Hamiltonian addressing only states with at most
  one orbital excitation (``e'') per lattice site.  We motivate this
  restriction immediately after Eq.~(8) in our theory overview, as
  soon as we begin a technical discussion of multi-body interactions
  (page 7, line 20):

  \red{We are interested in the ground and low-lying orbital
    eigenstates of several interacting atoms on a single lattice site,
    as such states can be readily prepared and coherently addressed in
    current experiments with ultracold atoms [34].}

  We acknowledge that the referee found the paragraph containing this
  sentence confusing, and have consequently rewritten the paragraph
  between Eqs.~(8) and (9) entirely (see point \ref{pt:confusion}).
  In any case, our actual results in section III (page 8) take the
  general form of Eq.~(9) (page 7), and make no restrictions on the
  number of orbital excitations per lattice site.  The expression in
  Eq.~(10) is merely a simplified restriction of our technical
  results, made (i) to highlight important features of the low-lying
  orbital excitations in our multi-body SU($N$) theory, and (ii) to
  facilitate comparison with optical lattice clock experiments.

  To clarify this point, we have made more explicit the fact that our
  actual results take the form of Eq.~(9), and added additional
  discussion of Eq.~(10).  Specifically, we have replaced the
  following text (page 7, line 48):

  \red{Our low-energy effective theory exhibits SU($N$)-symmetric
    multi-body interactions, such that the effective interaction
    Hamiltonian can be written in the form
    \begin{align}
      H_{\t{int}}^{\t{eff}} = \sum_{M=2}^{2I+1} H_M, \tag{9}
    \end{align}
    where $H_M$ is an $M$-body Hamiltonian, $I$ is the total nuclear
    spin of each atom (e.g.~$I=9/2$ for ${}^{87}$Sr), and the sum
    terminates at $2I+1$ because this is the largest number of atoms
    which may initially occupy a single lattice site.  Focusing on the
    low-lying orbital state excitations of this theory, we find that
    SU($N$) symmetry allows us to express multi-body Hamiltonians in
    the simple form}

  with

  \green{Our low-energy effective theory exhibits SU($N$)-symmetric
    multi-body interactions, such that the effective interaction
    Hamiltonian can be written in the form
    \begin{align}
      H_{\t{int}}^{\t{eff}} = \sum_{M=2}^{2I+1} \sum_{p\ge1} H_M^{(p)},
      \tag{9}
    \end{align}
    where $H_M^{(p)}$ is an $M$-body Hamiltonian of order $p$ in the
    coupling constants $G^{qr}_{st}$, $I$ is the total nuclear spin of
    each atom (e.g.~$I=9/2$ for ${}^{87}$Sr), and the sum terminates
    at $2I+1$ because this is the largest number of atoms which may
    initially occupy a single lattice site.  We explicitly compute all
    $M$-body Hamiltonians $H_M\equiv\sum_p H_M^{(p)}$ through order
    $p=3$, yielding effective two-, three-, and four-body
    interactions.  To facilitate understanding and comparison with
    experimental measurements of density-dependent atomic energy
    spectra [34], in which states with multiple orbital excitations on
    a single lattice site are spectroscopically off-resonant, we also
    consider a restriction of our theory to states with at most one
    orbital excitation per lattice site.  In this case, we find that
    SU($N$) symmetry allows us to express multi-body Hamiltonians in
    the simple form}

  We have also modified the following sentence in our summary (page
  28, line 43):

  \red{The SU($N$) symmetry of $M$-body Hamiltonians allowed us to
    express them in a simple form, and fully characterize their
    corresponding eigenstates and spectra.}

  to:

  \green{When considering the low-lying orbitalexcitations of our
    effective theory, the SU($N$) symmetry of $M$-body Hamiltonians
    allowed us to express them in a simple form, and to fully
    characterize their eigenstates and spectra.}

  Finally, the new appendix we have added as a part of point
  \ref{pt:description} (below) contains additional discussion of the
  restriction to at most one orbital excitation per lattice site.

  We hope that these changes clarify the scope of our work, in
  particular highlighting that the theory developed in section III
  applies to any number of orbital excitations, despite the
  restriction to the single-excitation case for the sake of comparison
  with experiments in section IV (page 18).

\item Concerning the next example:

  \blue{In addition, the restriction to a single occupation of the
    nuclear spin state is never motivated or justified but just
    introduced at some point.}

  \label{pt:single_nuclear}

  The referee is correct that we restrict our theory to the case of
  single occupation of nuclear spin states on any given lattice site.
  We do so because (i) this regime is precisely that relevant to
  current experiments with ultracold atoms, which begin with
  ground-state preparation of the atoms, and (ii) this restriction
  greatly simplifies our theory.  We explained both the motivation and
  the justification for this restriction in the overview of our theory
  (page 7, line 40):

  \red{As our work is motivated by progress in experiments which
    initialize all atoms in identical (i.e.~ground) orbital and
    motional states, to simplify our theory we assume that any $N$
    atoms on a single lattice site occupy $N$ nuclear spin states.
    Multiple occupation of a single nuclear spin state on the same
    lattice site is initially forbidden by fermionic statistics, and
    subsequently cannot be violated in the absence of tunneling and
    hyperfine coupling.}

  Deferring the referee's later comment that this text was confusing
  to point \ref{pt:confusion}, we emphasize the fact that our theory
  is the first to include hyperfine structure when considering
  multi-body interactions of ultracold atoms on a lattice.  Of course,
  a more general theory relaxing the above restriction (i.e.~on the
  occupation of nuclear spin states) will eventually be necessary for
  future experiments with a shallower lattice or additional control
  fields.  Such a general theory will need to consider inter-site
  effects or hyperfine coupling (or, less likely, non-ground-state
  preparation protocols), as these are the mechanisms by which a
  nuclear spin state can become multiply-occupied on a single lattice
  site.  Given the that these mechanisms (i) are largely irrelevant
  for the experiments that prompted this work, and (ii) introduce
  major complications into the study of SU($N$) multi-body
  interactions, we believe that it is appropriate to leave such
  considerations for future work.

  For clarification, we have modified the sentence on page 7, line 43
  from

  \red{Multiple occupation of a single nuclear spin state on the same
    lattice site is initially forbidden by fermionic statistics, and
    subsequently cannot be violated in the absence of tunneling and
    hyperfine coupling.}

  to

  \green{Ground-state preparation implies that multiple occupation of
    a single nuclear spin state on any given lattice site is initially
    forbidden by fermionic statistics, and subsequently cannot occur
    in the absence of inter-site effects or hyperfine coupling.}

  We hope that this modification, together with the modifications
  mentioned in point \ref{pt:general}, clarifies our justification for
  restricting to the case of single occupation of nuclear spin states.

\item Concerning the following comments:

  \blue{The referee understands, that these restrictions are justified
    for the comparison with the experiment in Ref.[34]. However, it
    should be clearly pointed, that the analysis is not general and
    adapted to the description to these experiments in the abstract
    and introduction.}

  \label{pt:restrictions}

  We thank the referee for pointing out the potential to mislead
  readers on the generality of our theory, and hope that the new
  version of our manuscript does a better job of calibrating the
  expectations of a reader with regard to restrictions and
  approximations that we make.  In particular, we hope that the
  current text makes clear to readers that we are developing a theory
  for the purpose of understanding current optical lattice clock
  experiments, and not to understand SU($N$) multi-body interactions
  in full generality.

\item Concerning the description of the experiment with which we
  compare:

  \blue{Furthermore, a short description of the experiment would be
    mandatory for an independent manuscript.}

  \label{pt:description}

  We thank the referee for pointing out that a description of relevant
  experimental procedures would make our manuscript more
  self-contained, and we recognize the value in doing so.  We have
  therefore added a new appendix to our manuscript, summarizing all
  relevant experimental procedures:

  \green{{\bf Appendix A:~Summary of experimental procedures}}

  \green{The work in this paper is closely tied to the experimental
    work in ref.~[34]; for quick reference, we here provide a short
    summary of the experimental procedures used therein.  The
    experiment begins by preparing a degenerate gas of $10^4$-$10^5$
    (fermionic) ${}^{87}$Sr atoms, in a uniform mixture of their ten
    nuclear spin states, at $\sim0.1$ of their fermi temperature (10
    to 20 nanokelvin) [8, 52].  This gas is loaded into a primitive
    cubic optical lattice at the ``magic wavelength'' for which both
    ground (${}^1S_0$) and first-excited (${}^3P_0$) electronic
    orbital states of the atoms experience the same lattice potential.
    Lattice depths along the principal axes of the lattice are
    comparable in magnitude, with a geometric mean that can be varied
    from 30 to 80 $E_{\t{R}}$, where
    $E_{\t{R}}\approx3.5\times2\pi~\t{kHz}$ is the lattice photon
    recoil energy of the atoms.  These lattice depths are sufficiently
    large as to neglect tunneling on the time scales relevant to the
    experiment.  Atoms are then addressed by an external (``clock'')
    laser with an ultranarrow (26 mHz) linewidth, detuned by $\Delta$
    from the single-atom ${}^1S_0-{}^3P_0$ transition frequency.  When
    $\Delta=0$, non-interacting atoms in singly-occupied lattice sites
    thus undergo Rabi oscillations.  After a chosen time, the
    experiment removes all ground-state (${}^1S_0$) atoms from the
    lattice and uses absorption imaging to identify the remaining
    excited-state (${}^3P_0$) atoms.  As discussed in section IV, a
    similar procedure with $\Delta\ne0$ leads to the spectrum shown in
    figure 3, thereby constituting a measurement of the many-body
    atomic excitation spectrum.  The many-body atomic excitation
    spectrum is different from the single-body atomic excitation
    spectrum due to the presence of inter-atomic interactions.  We
    note that this spectroscopic protocol addresses only
    singly-excited orbital states of any lattice site.  Doubly-excited
    states are manifestly off-resonant for single-photon absorption
    processes, while multi-photon absorption processes are always
    off-resonant due to (i) the non-linearity (on the scale of kHz) of
    on-site electronic orbital excitation energies (resulting from
    non-uniform coupling constants $G_X$), combined with (ii) the
    ultranarrow linewidth (on the scale of mHz) of the interrogation
    laser.}

  We have also added, at the end of the first paragraph in our
  introduction that discusses our technical contributions, a sentence
  directing readers to the above summary of experimental procedures.
  Specifically, the text (page 4, line 31)

  \red{... through the exquisite capabilities with OLCs.}

  now reads

  \green{... through the exquisite capabilities with OLCs.  We briefly
    summarize the experimental procedures relevant to our work in
    Appendix A.}

  We have likewise directed readers to the new appendix at the
  beginning of the experiment/theory comparisons in section IV (page
  18), such that the first sentence of this section, previously

  \red{Current experiments with ultracold ${}^{87}$Sr on a lattice can
    coherently address ground states and their low-lying orbital
    excitations for up to five atoms per lattice site [34].}

  now reads

### 2019-02-07 multi_body_interactions/papers/theory/paper.tex
updates/changes from responses to NJP referee 1

% general physics / math packages and commands
\usepackage{physics,amsmath,amssymb,braket,dsfont,bm}
\renewcommand{\t}{\text} % text in math mode
\newcommand{\f}{\dfrac} % shorthand for fractions
\newcommand{\p}[1]{\left(#1\right)} % parenthesis
\renewcommand{\sp}[1]{\left[#1\right]} % square parenthesis
\renewcommand{\set}[1]{\left\{#1\right\}} % curly parenthesis
\newcommand{\bk}{\braket} % shorthand for braket

\newcommand{\g}{\text{g}}
\newcommand{\e}{\text{e}}
\newcommand{\x}{\text{x}}
\newcommand{\y}{\text{y}}
\newcommand{\z}{\text{z}}

@@

\begin{abstract}
  Rapid advancements in the experimental capabilities with ultracold
  alkaline-earth-like atoms (AEAs) bring to a surprisingly near term
  the prospect of performing quantum simulations of spin models and
  lattice field theories exhibiting SU($N$) symmetry.  In order to
  analyze recent experiments preparing high density samples of
  strongly interacting ${}^{87}$Sr atoms in a three-dimensional
  optical lattice, we develop a low-energy effective theory of
  fermionic AEAs which exhibits emergent multi-body SU($N$)-symmetric
  interactions, where $N$ is the number of atomic nuclear spin levels.
  We fully characterize the low-lying excitations in our effective
  theory, and compare predictions of many-body interaction energies
  with direct measurements of many-body excitation spectra in an
  optical lattice clock.  Our work makes the first step in enabling a
  controlled, bottom-up experimental investigation of multi-body
  SU($N$) physics.
\end{abstract}

@@

In this work, we investigate the first experimental capabilities with
ultracold fermionic AEAs to prepare high-density samples in a 3-D
lattice with multiple occupation of individual lattice sites
\cite{goban2018emergence}.  Specifically, we consider ground-state
preparation of isolated few-body systems in the deep-lattice limit,
and carry out a bottom-up investigation of emergent multi-body
interactions on multiply-occupied lattice sites.  These multi-body
interactions appear in a low-energy effective theory of the atoms, and
inherit the SU($N$) symmetry of their bare, pair-wise interactions,
thereby enabling experimental studies of multi-body SU($N$) physics
through the exquisite capabilities with OLCs.  We briefly summarize
the experimental procedures relevant to our work in Appendix
\ref{sec:experiment_summary}.

Though effective multi-body interactions have previously been studied
in the context of harmonically \cite{johnson2012effective,
  yin2014universal} and lattice-confined \cite{johnson2009effective}
neutral bosons prepared in a single hyperfine state, our work deals
for the first time with fermions that have internal degrees of freedom
and multiple collisional parameters.  Nonetheless, due to the SU($N$)
symmetry of these collisions, we are able to find a simple way to
express effective multi-body Hamiltonians and fully characterize their
low-lying many-body energy eigenstates.  While some past work has
detected experimental signatures of multi-body interactions in the
form of quantum phase revivals \cite{will2010timeresolved}, we perform
a direct comparison of the many-body interaction energies predicted by
our low-energy effective theory with experimental measurements of
density-dependent atomic energy spectra performed in
ref.~\cite{goban2018emergence}, similarly to the measurements with
bosons performed in ref.~\cite{mark2011precision}.

The remainder of this paper is structured as follows.  In section
\ref{sec:overview} we provide an overview of the relevant one- and
two-body physics of ultracold atoms on a lattice, and summarize our
main result of effective multi-body interactions between these atoms.
In section \ref{sec:low_energy} we discuss our method for deriving a
low-energy effective theory, provide a perturbative expansion for the
net effective Hamiltonian, and compute all $M$-body Hamiltonians
through third order in the low-energy effective theory.  We then
analyze the low-lying excitations of the effective theory in section
\ref{sec:spectra}, comparing spectral predictions with experimental
measurements, and study the orbital-state dynamics of nuclear spin
mixtures interrogated via Rabi spectroscopy.  Finally, we summarize
and conclude our findings in section \ref{sec:summary}, and provide
some discussion of future outlooks.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overview}
\label{sec:overview}

Consider a collection of fermionic atoms with total nuclear spin $I$
and two atomic orbital states loaded into a translationally invariant
3-D optical lattice.  While an external trapping potential will
generally break discrete translational symmetry of the lattice, any
background inhomogeneity can be made negligible by spectroscopically
addressing a sufficiently small region of the lattice
\cite{goban2018emergence}.  Throughout this paper, we work strictly in
the deep-lattice regime with negligible tunneling between lattice
sites.  We also assume a so-called ``magic-wavelength'' lattice for
which both atomic orbital states experience identical lattice
potentials \cite{ye2008quantum}.  The single-particle Hamiltonian of
the atoms can then be written in the form
\begin{align}
  H_0 = \sum_{i,n,\mu,s} E_n \c_{in\mu s}^\dag \c_{in\mu s},
  \label{eq:H_0}
\end{align}
where $\c_{in\mu s}$ is a fermionic operator which annihilates a
single atom on lattice site $i\in\mathbb{Z}^3$ in motional state
$n\in\mathbb{N}_0^3$ with nuclear spin $\mu\in\set{-I,-I+1,\cdots,I}$
(i.e.~projected onto a quantization axis) and orbital state
$s\in\set{\g,\e}$; and $E_n$ is the energy of a single atom in
motional state $n$.  In a harmonic trap approximation we would have
$E_n=\p{3/2+n_\x+n_\y+n_\z}\omega$ for an on-site angular trap
frequency $\omega$ (with the reduced Planck constant $\hbar=1$
throughout this paper), but in general the aharmonicity of the lattice
potential will cause a non-negligible shift in motional state
energies.

>>>>

% general physics / math packages and commands
\usepackage{physics,amsmath,amssymb,braket,dsfont}
\renewcommand{\t}{\text} % text in math mode
\newcommand{\f}{\dfrac} % shorthand for fractions
\newcommand{\p}[1]{\left(#1\right)} % parenthesis
\renewcommand{\sp}[1]{\left[#1\right]} % square parenthesis
\renewcommand{\set}[1]{\left\{#1\right\}} % curly parenthesis
\newcommand{\bk}{\braket} % shorthand for braket

\renewcommand{\d}{\text{d}}
\newcommand{\g}{\text{g}}
\newcommand{\e}{\text{e}}
\newcommand{\x}{\text{x}}
\newcommand{\y}{\text{y}}
\newcommand{\z}{\text{z}}

@@

\begin{abstract}
  Rapid advancements in the experimental capabilities with ultracold
  alkaline-earth-like atoms (AEAs) bring to a surprisingly near term
  the prospect of performing quantum simulations of spin models and
  lattice field theories exhibiting SU($N$) symmetry.  Motivated in
  particular by recent experiments preparing high density samples of
  strongly interacting ${}^{87}$Sr atoms in a three-dimensional
  optical lattice, we develop a low-energy effective theory of
  fermionic AEAs which exhibits emergent multi-body SU($N$)-symmetric
  interactions, where $N$ is the number of atomic nuclear spin levels.
  Our theory is limited to the experimental regime of
  \begin{enumerate*}
  \item a deep lattice, with
  \item at most one atom occupying each nuclear spin state on any
    lattice site,
  \end{enumerate*}
  where the latter restriction is a consequence of initial
  ground-state preparation.  We fully characterize the low-lying
  excitations in our effective theory, and compare predictions of
  many-body interaction energies with direct measurements of many-body
  excitation spectra in an optical lattice clock.  Our work makes the
  first step in enabling a controlled, bottom-up experimental
  investigation of multi-body SU($N$) physics.
\end{abstract}

@@

In this work, we investigate the first experimental capabilities with
ultracold fermionic AEAs to prepare high-density samples in a 3-D
lattice with multiple occupation of individual lattice sites
\cite{goban2018emergence}.  Specifically, we consider ground-state
preparation of isolated few-body systems in the deep-lattice limit,
and carry out a bottom-up investigation of emergent multi-body
interactions on multiply-occupied lattice sites.  These multi-body
interactions appear in a low-energy effective theory of the atoms, and
inherit the SU($N$) symmetry of their bare, pair-wise interactions,
thereby enabling experimental studies of multi-body SU($N$) physics
through the exquisite capabilities with OLCs.  Our theory is limited
to the experimental regime of at most one atom occupying each nuclear
spin state on any lattice site, which is a consequence initial
ground-state preparation.

Though effective multi-body interactions have previously been studied
in the context of harmonically \cite{johnson2012effective,
  yin2014universal} and lattice-confined \cite{johnson2009effective}
neutral bosons prepared in a single hyperfine state, our work deals
for the first time with fermions that have internal degrees of freedom
and multiple collisional parameters.  We perform a direct comparison
between the many-body interaction energies predicted by our low-energy
effective theory and the experimental measurements of the
density-dependent orbital excitation spectra performed in
ref.~\cite{goban2018emergence}, similarly to the measurements with
bosons performed in ref.~\cite{mark2011precision}.  To facilitate this
comparison of excitation spectra and characterize the low-lying
excitations in our effective theory, we consider a restriction of the
theory to states with at most one orbital excitation per lattice site,
and find that the SU($N$) symmetry of atomic collisions allow
effective multi-body interactions to take a remarkably simple form.
We note that past work has also detected experimental signatures of
multi-body interactions in the form of quantum phase revivals
\cite{will2010timeresolved}.

The remainder of this paper is structured as follows.  In section
\ref{sec:overview} we summarize the experimental procedures relevant
to our work, provide an overview one- and two-body physics of
ultracold atoms on a lattice, and preview our main technical results.
In section \ref{sec:low_energy} we discuss our method for deriving a
low-energy effective theory, provide a perturbative expansion for the
net effective Hamiltonian, and compute all $M$-body Hamiltonians
through third order in the low-energy effective theory.  We then
analyze the low-lying excitations of the effective theory in section
\ref{sec:spectra}, comparing spectral predictions with experimental
measurements, and study the orbital-state dynamics of nuclear spin
mixtures interrogated via Rabi spectroscopy.  Finally, we summarize
and conclude our findings in section \ref{sec:summary}, and provide
some discussion of future outlooks.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background and overview}
\label{sec:overview}

The work in this paper is closely tied to the experimental work in
ref.~\cite{goban2018emergence}; we begin with a short summary of the
relevant experimental procedures therein.  The experiment begins by
preparing a degenerate gas of $10^4$-$10^5$ (fermionic) ${}^{87}$Sr
atoms in a uniform mixture of their ten nuclear spin states and at
$\sim0.1$ of their fermi temperature ($\sim10$ nanokelvin)
\cite{campbell2017fermidegenerate, marti2018imaging}.  This gas is
loaded into a primitive cubic optical lattice at the ``magic
wavelength'' for which both ground (${}^1S_0$) and first-excited
(${}^3P_0$) electronic orbital states of the atoms experience the same
lattice potential \cite{ye2008quantum}.  Lattice depths along the
principal axes of the lattice are roughly equal in magnitude, with a
geometric mean that can be varied from 30 to 80 $E_{\t{R}}$, where
$E_{\t{R}}\approx3.5\times2\pi~\t{kHz}$ is the lattice photon recoil
energy of the atoms (with the reduced Planck constant $\hbar=1$
throughout this paper).  These lattice depths are sufficiently large
as to neglect tunneling on the time scales relevant to the experiment.
The temperature of the atoms is also low enough to neglect thermal
occupation of states outside the ground-state manifold.

Once loaded into an optical lattice lattice, atoms are addressed by an
external (``clock'') interrogation laser with an ultranarrow (26 mHz)
linewidth, detuned by $\Delta$ from the single-atom ${}^1S_0-{}^3P_0$
transition frequency $\omega_0$.  After a fixed interrogation time,
the experiment turns off the interrogation laser, removes all
ground-state (${}^1S_0$) atoms from the lattice, and uses absorption
imaging to count any remaining excited-state (${}^3P_0$) atoms.
Non-interacting atoms in singly-occupied lattice sites undergo Rabi
oscillations when $\Delta=0$ during the interrogation time.  The
orbital excitation spectrum of multiply-occupied lattice sites,
however, is shifted by inter-atomic interactions, which results in
spectroscopic peaks (i.e.~local maxima in excited-state atom counts)
away from $\Delta=0$.  A sweep across different detunings $\Delta$ (on
the scale of inter-atomic interaction energies) thus constitutes a
measurement of the many-body orbital excitation spectrum.  We note
that this spectroscopic protocol addresses only singly-excited orbital
states of any lattice site.  Doubly-excited states are manifestly
off-resonant for single-photon absorption processes as
$\abs{\Delta}\ll\omega_0$, while multi-photon absorption processes are
always off-resonant due to
\begin{enumerate*}
\item the non-linearity ($\sim$kHz) of orbital excitation energies for
  a collection of interacting atoms, and
\item the ultranarrow linewidth ($\sim$mHz) of the interrogation
  laser.
\end{enumerate*}

Consider now a collection of fermionic atoms with total nuclear spin
$I$ and two atomic orbital states loaded into a translationally
invariant 3-D optical lattice.  While an external trapping potential
will generally break discrete translational symmetry of the lattice,
any background inhomogeneity can be made negligible by
spectroscopically addressing a sufficiently small region of the
lattice \cite{goban2018emergence}.  Throughout this paper, we work
strictly in the deep-lattice regime with negligible tunneling between
lattice sites, and assume that both atomic orbital states experience
identical lattice potentials.  The single-particle Hamiltonian of the
atoms can then be written in the form
\begin{align}
  H_0 = \sum_{i,n,\mu,s} E_n \c_{in\mu s}^\dag \c_{in\mu s},
  \label{eq:H_0}
\end{align}
where $\c_{in\mu s}$ is a fermionic operator which annihilates a
single atom on lattice site $i\in\mathbb{Z}^3$ in motional state
$n\in\mathbb{N}_0^3$ with nuclear spin $\mu\in\set{-I,-I+1,\cdots,I}$
(i.e.~projected onto a quantization axis) and orbital state
$s\in\set{\g,\e}$; and $E_n$ is the energy of a single atom in
motional state $n$.  In a harmonic trap approximation we would have
$E_n=\p{3/2+n_\x+n_\y+n_\z}\omega$ for an on-site angular trap
frequency $\omega$, but in general the aharmonicity of the lattice
potential will cause a non-negligible shift in motional state
energies.

In the absence of hyperfine coupling, as when addressing the spinless
${}^1S_0$ ($\g$) and ${}^3P_0$ ($\e$) orbital states of AEAs,
interactions between any two atoms are governed by their orbital
states alone, and are therefore characterized by four scattering
lengths $a_X$ with $X\in\set{\g\g,\e\g^-,\e\g^+,\e\e}$, where the $+$
($-$) superscript denotes symmetrization (anti-symmetrization) of a
two-body orbital state under particle exchange.  In the low-energy
limit, we can write the bare two-body interaction Hamiltonian in the
form \cite{gorshkov2010twoorbital}
\begin{align}
  H_{\t{int}} = \sum_{\substack{\mu<\nu\\s}}
  G_s \int \d^3x~ \hat\rho_{\mu s} \hat\rho_{\nu s}
  + G_+ \sum_{\mu,\nu}
  \int \d^3x~ \hat\rho_{\mu,\e} \hat\rho_{\nu,\g}
  + G_- \sum_{\mu,\nu} \int \d^3x~
  \hat\psi_{\mu,\e}^\dag \hat\psi_{\nu,\g}^\dag
  \hat\psi_{\nu,\e} \hat\psi_{\mu,\g},
  \label{eq:H_int_start}
\end{align}
where $\hat\psi_{\mu s}$ is a fermionic field operator for atoms with
nuclear spin $\mu$ and orbital state $s$;
$\hat\rho_{\mu s}\equiv\hat\psi_{\mu s}^\dag\hat\psi_{\mu s}$ is an
atomic density field operator; and the coupling constants $G_X$ are
defined in terms of the scattering lengths by
\begin{align}
  G_{s=\g,\e} \equiv \f{4\pi}{\mA}~ a_{ss}, &&
  G_\pm \equiv \f{2\pi}{\mA}~ \p{a_{\e\g+} \pm a_{\e\g-}},
  \label{eq:couplings}
\end{align}
where $\mA$ is the mass of a single atom.  Defining for brevity
\begin{align}
  G^{qr}_{st} \equiv \left\{
    \begin{array}{ll}
      G_q & ~ q = r = s = t \\
      G_+ & ~ q \ne r ~ \t{and} ~ (q,r) = (s,t) \\
      G_- & ~ q \ne r ~ \t{and} ~ (q,r) = (t,s) \\
      0 & ~ \t{otherwise}
    \end{array}\right.,
  \label{eq:coupling_tensor}
\end{align}
where $q,r,s,t\in\set{\g,\e}$ are orbital state indices, we can
alternately write the bare two-body interaction Hamiltonian in the
more compact form
\begin{align}
  H_{\t{int}}
  = \f12 \sum_{\substack{q,r,s,t\\\mu,\nu}} G^{qr}_{st} \int \d^3x~
  \hat\psi_{\mu s}^\dag \hat\psi_{\nu t}^\dag
  \hat\psi_{\nu r} \hat\psi_{\mu q}.
  \label{eq:H_int_fields}
\end{align}
For nuclear spins $\mu,\nu$, the symbol $G^{qr}_{st}$ gives the
coupling constant between the two-atom states
$(\mu,q)+(\nu,r)\leftrightarrow(\mu,s)+(\nu,t)$.
